---
title: 'Lab 3: Type your title here'
author: "Type your name here"
date: "Type the date here"
output:
  html_document:
    default: yes
    number_sections: yes
    theme: sandstone
  pdf_document:
    number_sections: yes
  word_document: default
---

This first section is meant to familiarize you with the format and syntax used in RMarkdown documents. Please go slowly through this section, comparing the text in the RMarkdown format to the output in the corresponding HTML file, to learn about the different components.

# Introduction to Markdown Syntax

We start a new section by using the pound sign, followed by the section title. The sections will be automatically numbered because of our header specifications.

## Use more pounds for deeper subsections

Start a new paragraph by simply adding a blank line.

> Make a blockquote out of text by preceding the lines
> with the "greater than" symbol.
>
>> You can also make nested blockquotes with more "greater than" signs.

Emphasize *text* (italics) by surrounding the text in single asterisks. Emphasize **more strongly** (bold) with double asterisks.

**Note**: When working with a RMarkdown report, it is a good idea to not use the console much, as the report window and the console window do not communicate directly with each other; actions you take in the console will not count when you "knit" your report. For that reason it is advisable to maximize the RMarkdown window by clicking the resizing icon on its upper-right corner. This will hide/minimize the console window.

- Make unnumbered lists by starting each item with a dash.
- Here's the second item!
    - You can create a sublist by starting the line with 4 spaces.
- **DO NOW**: Make the word "unnumbered" in the first bullet point appear in bold, and the word "second" in the second bullet point appear in italics.

1. Numbered lists use numbers
2. followed by periods.
3. **DO NOW**: Add a fourth item to this list.

Surrounding some text with `single backticks` makes it appear in "typewriter" font. This is useful for inline code. You probably won't need to use this feature much.

You can add horizontal rules by a sequence of three or more asterisks:

***********

They will be quite faint.

You can easily add subscripts~2~ or superscripts^3^.

There are many more specialized features that are available. A quick Google search can usually provide you with some resources, though the above should be sufficient for most reports.

***********

**DO NOW**: Replace the above ruler with a new subsection (two pound signs) with title "Introducing R Chunks".

R code chunks look like this (see below.) Note: the "backtick" key is typically found to the left of the number keys row.
```{r}
4 + 5
```
When the file compiles, the R code *chunks* will be evaluated and the results will be inserted into the document. That behavior can be changed (e.g. hide the results or prevent the code itself from appearing) by adding more in the `{r}` part. We will see how to do that in a little while.

You can insert a new code chunk by either typing the backticks and `{r}` yourself, or by using the `Insert` pulldown in the toolbar at the top of the RMarkdown window. Choose `R` from the pulldown and the appropriate code chunk will be inserted at your cursor.

**DO THIS NOW**: Insert a new code chunk and put a simple mathematical expression inside the chunk. Save and knit the document to see the added chunk.

In addition to compiling/knitting the document as a whole, you can execute individual chunks. There are buttons to the right of each chunk to help with that.

- The first button is a (rather dim) gear button for modifying chunk options. Use this button now to uncheck the "show warnings" option, and click apply. You will see the text `warning=FALSE` added to the top of the chunk.
- The second button re-executes all the chunks before this point. This is useful to do when you restart work on a particular report.
- The third button executes the current chunk and shows you the output. This is the most frequently used button. As you work on your report you will try out R commands in these chunks and run them to make sure they produce the output you want before moving on. Running these chunks takes the place of entering commands in the console.

DO NOW: Run the chunk you just created. You should see a brief output right below the chunk. You can leave it there for now or just remove it; it has no effect on the compiled report, but it does show you how these chunks might look in the compiled report. These "output" sections also have three convenient action buttons to their right. Take a moment to familiarize yourself with these buttons.

We will now begin our more specific statistical investigations.

# Statistical Work

Recall that before we start any work, in particular whenever we start a new report, we need to load our main functions package, `hanoverbase`. The following code chunk accomplishes that, and it should appear early on in the report.
```{r}
library(hanoverbase)
```
By default this chunk will print out lots of messages which we do not want to see. Use the gear button at the right side of the chunk to:

- Not show warnings, not show messages
- Show nothing, but run the code (option in the `Output` pulldown)

When you click `Apply`, this should add a couple of options to the top of your chunk, and when you compile/knit the report this chunk will not show up at all.

You should go ahead and **run your chunk now**. This will load the `hanoverbase` package, which we will use in the remainder.

We must next load the counties data. We similarly do not care about seeing the output of this chunk, so change its settings to match the previous chunk, and then run the chunk.
```{r}
data(counties)
```
To fully load the dataset we need to use the `View(counties)` command. BUT this command must be run in the Console, it cannot be in an R chunk. DO THIS NOW.

## Black Population across Indiana

Let us begin our investigations by considering the percent of Black population in the counties of Indiana. As a first step, we should create a new dataset containing just the counties in Indiana. Do this in the chunk below, by replacing the `"replace this"` part with working code. This should be very similar to the code we used in Lab 2 to pick out all the counties of Alaska.
```{r}
indianaCounties <- "replace this"
```
If you have done this correctly, then after you run the chunk you should see a new entry in the Environment pane, containing the `indianaCounties` data with 92 observations on 32 variables.

Type into this next chunk the code for producing the five number summary for the percent of black population in the Indiana counties dataset. Hint: Which column contains the relevant data?
```{r}
"Replace this"
```
If you have done this correctly, then when you run the chunk you should see the five number summary and the mean, standard deviation an count. You should be seeing a median of $0.65$ and a mean of $2.540217$. Notice that the mean is much larger than the median. What does that indicate about the underlying distribution?

We should next produce a suitable graph to visualize the distribution of this quantitative variable. Do so in this code chunk.
```{r}
"Replace this"
```
What are the main features of the distribution (range, modes, skewness, outliers)? Type your answer in this space, using the blockquote format. Make sure to express your answers in the context of this particular data. Use plain English as much as possible, as if explaining to a friend. (Do this in general for all such questions.)

> Answer here.
> You can use multiple lines but they all become a single paragraph.
> You can also write very long lines and they will wrap around. Pay attention to the line numbering on the left. You only need the ">" sign if you are truly starting a new line.
>
> Have a line with just the `>` sign if you want to create a new paragraph.

Don't forget to knit and look at how your answer appears in the HTML view.

Provide explanations for each of these features. E.g. how do the range and the mode make sense? Does the direction of the skewness, if any, make sense? How do you explain the outliers, if any (don't forget you can always examine the data table by double-clicking it in the Environment pane)?

> Use blockquote format for
> your answers here as well.

Which of the mean or median is a better measure of the center of the distribution? Explain why.

>

**IMPORTANT NOTE**: The code chunks that follow have a setting of ` eval=FALSE`, in order to allow you to compile/knit the rest of the file without errors. As you get to each chunk, you should remove the `eval=FALSE` parts, and the commas preceding them, leaving just a `{r}`.

At this point we may want to compare the individual county percentages with the overall state percentage of black population. We compute this percentage now, and along the way introduce two new commands.

The first command (`sum`) allows us to add up values across rows. In this instance, we want to compute the total number of blacks in the state of Indiana. We do this by computing the total number in a county by multiplying the black proportion with the county population, then add these all up:
```{r, eval=FALSE}
blackPopIndiana <- sum(~black / 100 * pop2010, data=indianaCounties)
```
When you run this chunk, the Environment pane should have a new entry for `blackPopIndiana`. Next we compute the overall population in Indiana, and combine the two to find the state percentage:
```{r, eval=FALSE}
popIndiana <- sum(~pop2010, data=indianaCounties)
blackPercentIndiana <- blackPopIndiana / popIndiana * 100
blackPercentIndiana
```
The third line simply prints out that value.

We now add that percent to our histogram of the black distribution (you did do a histogram earlier, right?).

```{r, eval=FALSE}
ladd(panel.abline(v=9.127))
```

This is a considerably more complicated command, but it basically says the following:

- We want to *add* to the previous graph (`ladd`)
- The shape we want to add is that of a line (`panel.abline`)
- It should be a **v**ertical line at a specific value (`v=`).

We will see later variations of this command for other graph addins.

How many counties in Indiana are above the state percentage? How is it possible that the state percentage is so high given that so few counties are above it?

> Write your answer here.

## Poverty across the States

We now investigate the percentages of people living below the poverty line across the various states (plus DC). Let us start by considering the poverty rates in individual counties.

Create a histogram of the poverty rates across all counties in the U.S. and describe the patterns of the distribution.
```{r}
"Replace this line with the appropriate command."
```

> Describe the distribution here.

What are the counties that stand out as having extremely high poverty rates? Can you find some plausible explanation for these high poverty rates (Google search encouraged). Make sure to look at the overall `counties` data, not just the Indiana counties.

> Type your answer here.

Compute the overall poverty rate in the U.S. in 2010. In order to do that, you need to copy and suitably modify the one-line and three-line chunks we used earlier to compute the black percentage in Indiana.

```{r}
"Replace this section with the appropriate commands."
```

What is the overall poverty rate? Where does this fall in your distribution (you can use `ladd` to add it to the graph if you wish)?

> Type your answer here.

To compare across states, a first step could be looking at the county percentages in a boxplot against the states:
```{r, fig.height=10}
bwplot(state~poverty, data=counties)
```

You might be tempted to use this plot to determine if a state has a higher level of poverty than another, but this could be misleading. For example, the plot might suggest that New Hampshire has a higher poverty rate than Rhode Island (e.g. it has a higher median/"middle box"). This is more easily seen if we filter only the counties from those two states:

```{r}
bwplot(state~poverty, data=counties %>% filter(state == "Rhode Island" | state == "New Hampshire"))
```

But in reality the overall poverty rate in Rhode Island is 12.19%, which is higher than New Hampshire's 7.83%. Explain how this can happen.

> Type your answer here.

Looking again at the first boxplot, we can see that Kentucky has a larger IQR/spread than Kansas. Explain what that means in practical terms.

> Type your answer here.

In the boxplot for Hawaii, there seem to be only 3 dots. What is going on? Does Hawaii have exactly three counties?

> Type your answer here.

We can examine and compare the IQRs across all states:
```{r, eval=FALSE}
povertyIqrs <- iqr(~poverty|state, data=counties)
```
Note the formula syntax here: `~poverty|state`. This should be read as: "work with the poverty values grouped by state". The method used in the front (`iqr` in this case) tells R what to do with these grouped values. We could have used for example `median` or `mean` or `sd` (standard deviation) instead.

In order to visualize the distribution of these iqrs, we sort them and then put them in a dotplot. We performed a similar computation using the pipe operator (`%>%`) to visualize the median female percentage by state in Lab 2. Do something similar in this instance (note the extra option of `fig.height=10`, this makes the graphs take up more vertical space than they normally would --- the default is 5):
```{r, fig.height=10}
"Replace this"
```

Which states have unusually *high* IQRs? What does that mean for these states?

> Type your answer here.

DC appears to have an IQR of 0. Does this tell us anything?

> Type your answer here.

Look back at the lab 3 instructions to find out how to submit your completed report. Don't forget to click the red button to close your RStudio session, close all windows and log out, before you leave the lab.
